### 2020-05-23 Manuscript/manuscript.tex
add initial North China neighbor text

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results (Evans et al., this volume). While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland. The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}. Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age. For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions. However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are well-constrained (Table 2). An additional not-rated pole included in the present compilation is the new pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group as presented in \cite{Eyster2019a} are also included.

@@

Following the amalgamation of the Archean provinces in Laurentia ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging given a multi-state tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

@@

%\subsection{Evaluating Laurentia's paleogeographic neighbors}
%
%\subsubsection{Baltica}
%
%\subsubsection{North China}
%
%\subsubsection{Australia}
%
%\subsubsection{Kalahari}
% U-Pb zircon ages from the portion of the Kalahari craton now exposed in Dronning Maud Land, East Antarctica have been interpreted to indicate continental arc magmatism between ca. 1160 to
%1070 Ma with a a peak of activity ca. 1110–1090 Ma  \citep{Wang2020a}. These age constraints as well as additional geochemical data have been interpreted to indicate subduction under Kalahari in the late Mesoproterozoic.
% This magmatic activity was followed by amphibolite to granulite facies metamorphism between ca. 1080 and 1030 Ma \citep{Wang2020a}.

>>>>

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications (Fig. \ref{fig:Laurentia_poles} and Table 2). The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results (Evans et al., this volume). While many factors associated with paleomagnetic poles can be assessed quantitatively through Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland. The `A' rating refers to poles that are judged to be of such high quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. The `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2014a}. Many of these poles in the Paleomagia database are quite valuable for reconstruction and should not be dismissed from being considered in paleogeographic reconstructions. However, there are ambiguities associated with many of the poles not given Nordic `A' or `B' ratings in terms of how well the nature of the remanence is constrained, including its age. For example, there are rich data associated with intrusive and metamorphic lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than the ages of remanence associated with dated eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating with the exception of the `B' rated pole from the ca. 1015 Ma Haliburton intrusions. However, while any one of these Grenville poles could be interpreted to suffer from large temporal uncertainty, the overall preponderance of poles in a similar location at the time suggests that they need to be taken seriously within paleogeographic reconstructions of Laurentia (although an alternative view of an allochthonous origin put forward by \citet{Halls2015a} is discussed below). In this compilation, the poles of \cite{Brown2012a} from the Adirondack highlands are used wherein the magnetic mineralogy and associated relative ages of remanence are relatively well-constrained (Table 2). An additional not-rated pole included in the present compilation is the new pole for the ca. 1144 Ma Ontario lamprophyre dikes \citep{Piispa2018a} that strengthens the position of Laurentia at the time and coincides with the position of the poles from the ca. 1140 Ma Abitibi dikes \citep{Ernst1993a}. This pole will likely receive an `A' rating when assessed at the next Nordic paleomagnetism workshop. Poles from the Neoproterozoic Chuar Group of southerwest Laurentia as presented in \cite{Eyster2019a} are also included.

@@

Following the amalgamation of the Archean provinces in Laurentia ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained, but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging given a multi-stage tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

@@

%\subsection{Evaluating Laurentia's paleogeographic neighbors}
%%
%%\subsubsection{Baltica}
%%
%\subsubsection{North China}
%
%The latest Mesoproterozoic to earliest Neoproterozoic pole path of the North China craton includes a swath of paleomagnetic poles with a similar arc length to the Keweenawan Track to Grenville Loop of Laurentia's APWP (\citealp{Zhao2019a}; Zhang et al., this volume). While the chronostratigraphic age constraints on these North China poles are much looser than those from Laurentia, \cite{Zhao2019a} propose that the North China poles can be aligned with the Keweenawan Track to reconstruct North China as being conjoined to the northwest margin of Laurentia from prior to ca. 1110 Ma into the early Neoproterozoic. North China would have been at polar latitudes ca. 1110 Ma and moved rapidly with Laurentia as it transited towards the equator. \cite{Zhao2019a} also argued that similarity in the detrital zircon age spectra between early Neoproterozoic sediments in NW Laurentia and North China basins supports this reconstruction. In particular, sediment transport from Laurentia could provide a source for ca. 1.18 Ga zircons (from the Shawinigan orogen) and ca. 1.08 Ga zircons (from the Grenville orogen). In the Laurentia basins, ca. 1.6 Ga zircons without a clear Laurentia source could be sourced from North China craton granites (e.g. \citealp{Wang2020a}). If North China was in this position, the timing of its arrival of North China is unclear. The ca. 1220 Ma dikes pole of the North China craton is not coincident with the ca. 1237 Ma Sudbury dikes pole in this reconstructed position leading \cite{Zhao2019a} and Zhang et al. (this volume) to suggest that North China arrived on the Laurentian margin between ca. 1220 and 1110 Ma although they note a lack of evidence for known North China orogenesis at this time. In terms of departing from this position, one possibility is that

%CHECK THE SUDBURY POINT FOR MYSELF

%
%\subsubsection{Australia}
%
%\subsubsection{Kalahari}
% U-Pb zircon ages from the portion of the Kalahari craton now exposed in Dronning Maud Land, East Antarctica have been interpreted to indicate continental arc magmatism between ca. 1160 to
%1070 Ma with a a peak of activity ca. 1110–1090 Ma  \citep{Wang2020a}. These age constraints as well as additional geochemical data have been interpreted to indicate subduction under Kalahari in the late Mesoproterozoic.
% This magmatic activity was followed by amphibolite to granulite facies metamorphism between ca. 1080 and 1030 Ma \citep{Wang2020a}.
